### 2017-06-22 paper/lexicographic-programs.tex
changed definition of pCFG to allow for multiple initial valuations

We now discuss how to leverage the mathematical results of the previous section to provide a sound proof rule for almost-sure termination of probabilistic programs. Hence, for the rest of this section we fix a \PP{} $\program$ and the associated pCFG $\pCFG_\program=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$.

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that $\preexp{\lem_j}(\cfg{\sigma}{i}(\run)) \leq 
\lem_j(\cfg{\sigma}{i}(\run))-\eps$ (which entails 
$\preexp{\lem_{j'}}(\cfg{\sigma}{i}(\run)) =
\lem_{j'}(\cfg{\sigma}{i}(\run))-\eps$ for all $1\leq j'< j$). Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
that irrespective of the initial choice of $\sigma$ it holds, for each $1\leq j 
\leq n$, that $\E^\sigma[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq\preexp{\lem_{j}}(\cfg{\sigma}{}) $. This can be achieved by a 
standard~\cite{xxx}, through somewhat technical computation, which we defer to 
\AppendixMaterial.

Since  $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
follows that $\probm^{\sigma}\ttime <\infty) =1$, irrespective of scheduler 
$\sigma$.
\end{proof}

>>>>

We now discuss how to leverage the mathematical results of the previous section 
to provide a sound proof rule for almost-sure termination of probabilistic 
programs. Hence, for the rest of this section we fix a \PP{} $\program$ and the 
associated pCFG 
$\pCFG_\program=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$.

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that $\preexp{\lem_j}(\cfg{\sigma}{i}(\run)) \leq 
\lem_j(\cfg{\sigma}{i}(\run))-\eps$ (which entails 
$\preexp{\lem_{j'}}(\cfg{\sigma}{i}(\run)) =
\lem_{j'}(\cfg{\sigma}{i}(\run))-\eps$ for all $1\leq j'< j$). Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq\preexp{\lem_{j}}(\cfg{\sigma}{}) $. This can be achieved by a 
standard~\cite{xxx}, through somewhat technical computation, which we defer to 
\AppendixMaterial.

Since  $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
follows that $\probm^{\sigma}_{\vecinit}\ttime <\infty) =1$, irrespective of 
$\sigma$ and $\vecinit$.
\end{proof}
